Uncertainty in the evolution of social learning

### 2022-05-08 ms.tex
added some additional paragraphs to analysis

\newcommand{\pisub}[1]{\pi_{\mathrm{#1}}}
\newcommand{\pilow}{\pisub{low}}
\newcommand{\pihigh}{\pisub{high}}

@@

We ran the simulation for $T=1000$ time steps. At
the end of each generation agents reproduce and then die off. 
Those selected to reproduce
pass on their boolean social learning trait, $s$, without mutation,
which specifes whether agents learn socially. We developed a series of
computational analyses where we systematically vary the uncertainty
variables and observe the average prevalence of social learning trait $s$
across agents and trials, which is our primary outcome measure, 
denoted $\langle s \rangle$.

@@

We manipulated environmental uncertainty parameters described above, $u$,
$\pisub{low}$, $B$, and $M$, to examine their effects on our
main outcome measure, the prevalence of social learning $s$. For each parameter setting
in our analysis we calculated the average value of $s$ at the final time step
of $T=1000$ across 100 runs for each combination of uncertainty parameter values.

To analyze the effect of various forms of uncertainty on social learning 
prevalence in our Analysis, we initialized several model runs with
different uncertainty parameters $u$, $\pisub{\mathrm{low}}$, $B$,
and $M$. We varied $u \in \{0.0, 0.1, \ldots, 1.0\}$ for each combination of
the following parameters:
$\pisub{low} \in \{0.1, 0.5, 0.8\}$, $B \in \{2, 4, 10\}$, $M \in \{B/2, B, 2B\}$.

@@

For social learning to evolve, it must improve upon individual learning. 
We found that different parameter settings indeed change how advantageous social learning
is compared to individual learning (Figure~\ref{fig:meanNetPayoffs}). 
To analyze the effect of uncertainty on
payoffs, it is useful to identify some constant reference values which we will
use to evaluate the benefit of social learning. First, the optimal expected payoff
is the optimal behavior payoff times the lifespan, which we can denote
$\langle \pi^*(L) \rangle = L\cdot\pihigh$. The expected base payoff by chance is 
$\langle \pi_{\text{base}} \rangle = L \left[ (B - 1) \pilow + \pihigh \right]$, which
an agent would receive on average from performing purely random behaviors. 
Most importantly for understanding the benefit and evolution of social learning,
we calculated by simulation the expected payoff to individual learners for
each triplet ($\pilow$, $B$, $L$), which we denote $\langle \pi_I \rangle$
and plot for reference in Figure~\ref{fig:meanNetPayoffs}. Note $\langle \pi_I
\rangle$ is insensitive to $u$, which only affects the benefit of social
learning.

We observe across parameter settings that social learning outperforms
individual learning when $u$ is small. In fact, when $u=0.0$, social learning 
enables each new generation to know the optimal behavior and achieve the maximum
expected payoff for their lifespan, $\langle \pi^*(L) \rangle$. 
However, in some cases, even when $u$ is small, 
the returns on social learning are truly marginal due to 
relatively high, uniform expected payoffs whether or not agents identify the optimal
behavior, especially when $\pilow = 0.8$. In intermediate cases, examining the decay
of the benefit of social learning informs our understanding of social learning
evolution analyzed above.  As our analysis of social learning evolution suggested,
social learning is only beneficial up to a point. When $\pilow=0.1$, increasing $B$
led both to larger differences between

\subsection{Evolutionary outcomes vary in their certainty}

Finally, we found that the \emph{evolutionary uncertainty} of the system increases
around the values of $u$ where $\meansl$ goes from 1 to 0. 
In one sense
To confirm and expand our analysis of this uncertainty, we 
also measured environmental uncertainty in terms of the
expected number of time steps to convergence, $\langle T \rangle$. Evolutionary
uncertainty is an outcome measure---an emergent form of uncertainty due to
weak selection pressures. When 
$\meansl \approx 0.5$ this means that approximately equal numbers of trials
ended with all agents as social learners as they ended with all agents 
individual learners. We found that increased $\pilow$
tended to increase the overall uncertainty in terms of $\langle T \rangle$.
Increasing $B$ also increased the $u$ at which $\langle T \rangle$ peaked.
This indicates the evolutionary system is significantly less certain of 
which behavior is optimal around critical uncertainty parameter ensembles.

@@

\begin{table}[h]
  \caption{Max. iterations = 5000.}
  \centering
  \begin{tabular}{cccc} \toprule
    $B$ & \# not at fixation & \# total series & Pct. not fixated \\
    \midrule  
    2  & 0  & 132000 & 0.0 \% \\
    4  & 0  & 132000 & 0.0 \% \\
    10 & 44 & 132000 & 0.00033  \% \\
    \bottomrule
  \end{tabular} 
\end{table}

>>>>

\newcommand{\pisub}[1]{\pi_{\mathrm{#1}}}
\newcommand{\pilow}{\pisub{low}}
\newcommand{\pihigh}{\pisub{high}}
\newcommand{\piI}{\langle \pisub{I} \rangle}
\newcommand{\piS}{\langle \pisub{S} \rangle}

@@

We ran the simulation for a maximum of $T_{max}=20000$ time steps. If after a
generation all agents are either social or individual learners then the
simulation is stopped since there is no mutation. At
the end of each generation agents reproduce and then die off. 
Those selected to reproduce
pass on their boolean social learning trait, $s$, without mutation,
which specifes whether agents learn socially. We developed a series of
computational analyses where we systematically vary the uncertainty
variables and observe the average prevalence of social learning trait $s$
across agents and trials, which is our primary outcome measure, 
denoted $\langle s \rangle$.

@@

We manipulated environmental uncertainty parameters described above, $u$,
$\pisub{low}$, $B$, and $M$, to examine their effects on our
main outcome measure, the prevalence of social learning $s$. For each parameter setting
in our analysis we calculated the average value of $s$ at the final time step,
either at the maximum $T_{\text{max}}=20000$ across 1000 trials 
for each combination of uncertainty parameter values.

To analyze the effect of various forms of uncertainty on social learning 
prevalence in our Analysis, we initialized several model runs with
different uncertainty parameters $u$, $\pisub{\mathrm{low}}$, $B$,
and $M$. We varied $u \in \{0.0, 0.1, \ldots, 1.0\}$ for each combination of
the following parameters:
$\pisub{low} \in \{0.1, 0.45, 0.8\}$, $B \in \{2, 4, 10\}$, $L \in \{1, B/2, B, 2B\}$.

@@

For social learning to reliably evolve, it must be more beneficial
(in terms of payoffs) than individual learning. Inversely, social learning
should be suppressed when it misleads agents away from quickly identifying the
optimal behavior through individual learning.
We found that different parameter settings indeed change how 
advantageous social learning
is compared to individual learning (Figure~\ref{fig:meanNetPayoffs}).
We find that the transition from $\meansl = 1$ to $\meansl = 0$ occurs when
the mean net payoff among populations of all social learners is close to 
that of the expected individual learner payoff, $\piI$.

To analyze the effect of uncertainty on
payoffs, it is useful to identify some constant reference values which we will
use to evaluate the benefit of social learning. First, the optimal expected payoff
is the optimal behavior payoff times the lifespan, which we can denote
$\langle \pi^*(L) \rangle = L\cdot\pihigh$. The expected base payoff by chance is 
$\langle \pi_{\text{base}} \rangle = L \left[ (B - 1) \pilow + \pihigh \right]$, which
an agent would receive on average from performing purely random behaviors. 
Most importantly for understanding the benefit and evolution of social learning,
we calculated by simulation the expected payoff to individual learners for
each triplet ($\pilow$, $B$, $L$), which we denote $\piI$
and plot for reference in Figure~\ref{fig:meanNetPayoffs}. Note 
$\piI$ is insensitive to $u$, which only affects the benefit of social
learning.

We observe across parameter settings that social learning outperforms
individual learning when $u$ is small. In fact, when $u=0.0$, social learning
enables each new generation to know the optimal behavior and achieve the
maximum expected payoff for their lifespan, $\langle \pi^*(L) \rangle$.
However, in some cases, even when $u$ is small, the returns on social learning
are truly marginal due to relatively high, uniform expected payoffs whether or
not agents identify the optimal behavior, especially when $\pilow = 0.8$. In
intermediate cases, examining the decay of the benefit of social learning
informs our understanding of social learning evolution analyzed above.

As our analysis of social learning evolution suggested, social learning is only
beneficial up to a point. When $\pilow=0.1$, increased $B$ led both to larger
differences between payoffs when social learning was more beneficial, and it
led to $\piS > \piI$ across more values of $u$. This occurs because when there
are fewer behaviors it is easier for non-social learners to identify the
optimal behavior on their own. Even relatively small values of $u$ disrupt
the fidelity of social learning to the point that it being misled is
prohibitively costly compared to simply finding the optimal on one's own.
A relatively short lifespan can also extend the utility of social learning
across more values of $u$ since there are fewer opportunities for agents to
independently identify and perform the optimal solution.

\subsection{Evolutionary outcome uncertainty}

Finally, we found that the \emph{evolutionary uncertainty} of the system
increases around the values of $u$ where $\meansl$ goes from 1 to 0.  We know
that outcomes are uncertain in this region because we know that $\meansl$
is equivalent to the fraction of outcomes that ended with all agents being
social learners. $1 - \meansl$ is the fraction of trials that ended with 
all agents becoming individual learners. Therefore, when $\meansl \approx 0.5$,
this means roughly half of the simulated trials fixated at $\meansl = 1.0$
while the other half fixated at $\meansl = 0.0$.
Only 44 of nearly 400k trials did not go to fixation ($\meansl = 1$ or $\meansl = 0$) 
within 20k maximum iterations (Table~\ref{tab:convergence}); all that did not
reach fixation had $B=10$.

We also measured environmental uncertainty more directly, in terms of the
expected number of time steps to convergence, $\langle T \rangle$.
Evolutionary uncertainty is thus an outcome measure---an emergent form of
uncertainty due to weak selection pressures. We found that increased $\pilow$
tended to increase the overall uncertainty in terms of $\langle T \rangle$.
Increasing $B$ also increased the $u$ at which $\langle T \rangle$ peaked.
This indicates the evolutionary system is significantly less certain of 
which behavior is optimal around critical uncertainty parameter ensembles.

@@

\begin{table}[h]
  \caption{Max. iterations = 5000.}
  \label{tab:convergence}
  \centering
  \begin{tabular}{cccc} \toprule
    $B$ & \# not at fixation & \# total series & Pct. not fixated \\
    \midrule  
    2  & 0  & 132000 & 0.0 \% \\
    4  & 0  & 132000 & 0.0 \% \\
    10 & 44 & 132000 & 0.00033  \% \\
    \bottomrule
  \end{tabular} 
\end{table}
